---
title: "corona"
author: ipi
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    number_sections: true
    code_folding: "hide"
---

```{r setuptransform, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

check.packages <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg))
    install.packages(new.pkg, dependencies = TRUE,
                     repos = "https://cran.rstudio.com")
  sapply(pkg, require, character.only = TRUE)
}

pkgList <- c("RCurl","WDI","readxl","httr","knitr","tidyverse",
             "estimatr","stargazer","countrycode","dataverse",
             "zip", "lubridate", "rvest", "fastDummies", "lfe")

check.packages(pkgList)

```


## Merge


```{r, message = FALSE, warning = FALSE}
wdi_region <- WDI::WDI_data$country %>% 
  data.frame() %>% 
  select(iso3c, region) %>%
  rename("country_id" = "iso3c")

df_wide <- read.csv("saved/corona_df.csv") %>% 
  select(country_id, country) %>% 
  distinct(country_id, country) %>% 
  arrange(country_id) %>% 
  left_join(wdi_region, by = "country_id")

to_join <- c("wdi_df.csv", 
             "pop_below14_df.csv", 
             "polity4_df.csv", 
             "swiid_df.csv", "eps_df.csv", "vdem_indexes_df.csv", 
             "polarization_df.csv", "trust_people_df.csv", 
             "trust_gov_df.csv", "populism_df.csv", "federalism_df.csv", 
             "dpi_checks_df.csv", "electoral_pressure_df.csv", 
             "gov_position_df.csv", "woman_leader_df.csv", 
             "past_epidemics_df.csv", "median_age_df.csv", 
             "qog_df.csv", "share_health_ins.csv", 
             "pandemic_prep_df.csv", "pop_density_df.csv", 
             "life_exp_df.csv", "respiratory_disease_df.csv", 
             "ghsi_detect_df.csv", "ghsi_df.csv", "ghsi_health_df.csv", 
             "ghsi_respond_df.csv","state_fragility_df.csv","pr_df.csv") 

  print(c(nrow(df_wide)))

  for(j in to_join){
    df_wide <- left_join(df_wide, read.csv(paste0("saved/", j)), by = c("country_id"))
  } 
  
  df_wide <- 
    df_wide   %>%
    mutate(share_older = 100*(older_m + older_f)/pop_tot, 
         pop_tot = pop_tot/1000000,
         pop_tot_log = log(pop_tot),
         gdp_pc = gdp_pc/1000,
         air_travel = log(1+air_travel),
         pop_density_log = log(pop_density),
         infection = ifelse(is.na(infection), 0, infection)
         )  
    
write.csv(df_wide, "saved/df_wide.csv")

print(dim(df_wide))
```

## Add time varying data

```{r}
corona    <- 
  read.csv("saved/corona_df.csv")%>% mutate(date = as.Date(date))

mobility  <- 
  read.csv("saved/mobility_level_df.csv") %>% mutate(date = as.Date(date))

weatherDF <- 
  read.csv("saved/weather_df.csv")

policyDF  <- 
  read.csv("saved/acaps_df.csv") %>% mutate(date = as.Date(date))

stringency  <- 
  read.csv("saved/stringency_df.csv") %>% mutate(date = as.Date(date))

df <- left_join(corona, select(df_wide, -country), by = c("country_id")) %>%
  left_join(policyDF, by = c("country_id","date")) %>%
  left_join(mobility, by = c("country_id","date")) %>%
  left_join(stringency, by = c("country_id","date")) %>% 
  left_join(weatherDF, by = c("country_id",  "year", "month"))   %>%
  mutate(deaths_cum_per_million = deaths_cum/pop_tot,
         deaths_cum_per_million_log = log(1 + deaths_cum_per_million),
  )  %>% 
  
  rename(geoid2 = country_id)

```


## Combine timelines

We make a version of the data that records number of deaths per million $T$ days after first hitting $D$ cases. Note currently data sparse so $T$ and $D$ low, and start benchmarked to cases rather than deaths.

Create cumul deaths variable lagged to days since $D$ cases.

```{r, message = FALSE, eval = TRUE}
D <- 10
T <- 10

df <- df %>%
  group_by(forcats::fct_explicit_na(geoid2)) %>%
  mutate(relative_start =
           ifelse(any(cases_cum >= D),
                  min(elapsed[cases_cum >= D]), NA),
         elapsed_rel = elapsed - relative_start) %>%
  ungroup() %>%
  group_by(forcats::fct_explicit_na(geoid2)) %>%
  mutate(
    relative_start_d = ifelse(any(deaths_cum >= D),
                              min(elapsed[deaths_cum >= D]), NA),
    elapsed_rel_d = elapsed - relative_start_d
  ) %>%
  ungroup()

rm(D,T)

```

## Country name

```{r}
# Generate country name
df <- df %>% 
  mutate(country = countrycode(geoid2,
                               origin = "iso3c",
                               destination = "country.name")) %>% 
  mutate(country = if_else(geoid2=="RKS", "Kosovo", country))

```

## Fill some missing variables down


## Stringency index needs to be filled down (b/c missings)

```{r}
df <- df %>% 
  arrange(geoid2, date) %>% 
  group_by(geoid2) %>% 
  fill(mobility_index, .direction = 'down')
```


# Clean duplicate variables resulted from merging
```{r}
df <- df %>% 
  dplyr::select(-starts_with("X."))
```