---
title: "Dashboard: Social and political correlates of Covid-19"
author: "WZB | Institutions and Political Inequality Group, `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    code_folding: "hide"
---

```{r packages and  load main data setup, include=FALSE}
rm(list = ls())
dographs <- TRUE

knitr::opts_chunk$set(echo = TRUE)

check.packages <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg))
    install.packages(new.pkg, dependencies = TRUE,
                     repos = "https://cran.rstudio.com")
  sapply(pkg, require, character.only = TRUE)
}

pkgList <- c("RCurl","WDI","readxl","httr","knitr","tidyverse", "pscl",
             "estimatr","stargazer","countrycode","dataverse",
             "zip", "lubridate", "rvest", "fastDummies", "lfe",
             "directlabels")
check.packages(pkgList)

## Load data 

df <- read.csv("saved/df.csv")
df_full <- read.csv("docs/df_full.csv")

## Veto players missingess issue

df <- df %>% 
  mutate(checks_veto = ifelse(checks_veto < 0, NA, checks_veto))
df_full <- df_full %>% 
  mutate(checks_veto = ifelse(checks_veto < 0, NA, checks_veto))

## Load measures 

measures <- 
  read.csv("measures.csv") %>% 
  filter(include == 1)
```


```{r source functions, eval=T, message=F, warning=F, echo=F}
source("code/thirds_graph.R")
source("code/add_linebreak.R")
source("code/results_list.R")
source("code/plotresults.R")
source("code/oos_plot.R")
source("code/short_table.R")
source("code/results_summary.R")
source("code/results_list_wd_families.R")
```



```{r get most recent date, include=F}
# In case todays data is less complete than yesterdays, take yesterdays
data_date <- max(as.Date(df$date_rep), na.rm = TRUE)
if( (df %>% filter(date_rep == data_date) %>% nrow()) <
        (df %>% filter(date_rep == data_date-1) %>% nrow())) data_date <- data_date - 1
df_today  <- df %>% filter(as.Date(date_rep) == data_date)
```

```{r get lasso results and def controls, include = F}

# controls <- measures[measures$family=="controls", ]$vars %>% as.character
# controls_labels <- measures[measures$family=="controls", ]$labels %>% as.character

## Load all Lasso results 
n_controls <- 4

lasso_res <- read_rds("saved/lasso_res.rds")

## Control_use 

controls_use <- lasso_res$lcoef_controls %>%
  filter(abs(coef)>0) %>%
  group_by(outcome) %>%
  filter(!duplicated(vars)) %>%
  filter(rank(-abs(coef))<=n_controls) %>%
  ungroup() %>%
  dplyr::select(vars, outcome, labels)

#### Health data quality will always be added to controls ####
outcomes <- c('deaths_cum_log',
              'deaths_cum_per_million_log',
              'stringency', 'mobility_index',
              'lockdown_bin', 'distancing_bin')


add_data_quality <- expand.grid(outcome = outcomes, vars = 'detect_index')
controls_use <- controls_use %>% 
  bind_rows(add_data_quality) %>% 
  mutate(labels = ifelse(vars == 'detect_index', 'Health data quality', labels))

## Controls for log total deaths

controls <- controls_use %>% filter(outcome == 'deaths_cum_log') %>% 
  pull(vars)
controls_labels <- controls_use %>% filter(outcome == 'deaths_cum_log') %>% 
  pull(labels)

## Control labels for the text 

controls_text <- paste(paste(tolower(controls_labels)[1:4], collapse = ", "), 
                       ", and ", 
                       tolower(controls_labels)[5])
controls_text <- gsub("\n", " ", controls_text)
controls_text <- gsub("ghsi", "GHSI", controls_text)
controls_text <- gsub("gov't", "government", controls_text)
controls_text <- gsub("imhe", "IMHE", controls_text)
controls_text <- gsub(" ,", ",", controls_text)

## Controls for log total deaths

controls_deaths_capita <- controls_use %>% 
  filter(outcome == 'deaths_cum_per_million_log') %>% 
  pull(vars)
controls_deaths_capita_labels <- controls_use %>% 
  filter(outcome == 'deaths_cum_per_million_log') %>% 
  pull(labels)

## Control labels for the text 

controls_deaths_capita_text <- paste(paste(tolower(controls_deaths_capita_labels)[1:4], collapse = ", "), 
                       ", and ", 
                       tolower(controls_deaths_capita_labels)[5])
controls_deaths_capita_text <- gsub("\n", " ", controls_deaths_capita_text)
controls_deaths_capita_text <- gsub("ghsi", "GHSI", controls_deaths_capita_text)
controls_deaths_capita_text <- gsub("gov't", "government", controls_deaths_capita_text)
controls_deaths_capita_text <- gsub("imhe", "IMHE", controls_deaths_capita_text)
controls_deaths_capita_text <- gsub(" ,", ",", controls_deaths_capita_text)
```

```{r functions, warning = FALSE, message = FALSE}

major_missing_because_of_controls <- 
  filter(df, !complete.cases(select(df, all_of(controls)))) %>% 
  filter(pop_tot>1 & !is.na(country) & country != "NA")

dropped <- major_missing_because_of_controls$country %>% 
  as.character %>% 
  unique 

major_missing_because_of_controls_pc <- 
  filter(df, !complete.cases(select(df, all_of(controls_deaths_capita)))) %>%
  filter(pop_tot>1 & !is.na(country) & country != "NA")

dropped_pc <- major_missing_because_of_controls_pc$country %>% 
  as.character %>% 
  unique 

results <- function(Y, 
                    X = vars, 
                    Xlab = NA,
                    XX = controls,
                    range = NULL, 
                    data = filter(df, geoid2 != "CN"), 
                    bivariate = FALSE, 
                    relative_dates = FALSE,
                    standardize = FALSE, 
                    drop_missing_c = TRUE,
                    model = "lm",
                    dayfreq = 5){

  if(drop_missing_c) data <- 
        filter(data, complete.cases(select(data, all_of(XX))))

  if(!relative_dates)  
    data$points <- data$elapsed
  
  if(relative_dates) {
    data$points <- data$elapsed_rel
    data <- filter(data, elapsed_rel < max(data$elapsed, na.rm = TRUE) - 80)
    }

  if(is.null(range)) 
    range <- c(ifelse(relative_dates, 0, 60), max(data$points, na.rm = TRUE))
  
  ## If standardize
  if (standardize)
    data <- data %>% 
      mutate_at(vars(one_of(X)),
                list(~./sd(., na.rm = T)))

  out <-

    lapply(seq(range[1], range[2], dayfreq), function(j) {

      lapply(X, function(x) {
      
        ## Formula
        rhs <- ifelse(bivariate, x, paste(unique(c(x, XX)), collapse = " + "))

        if (model == "lm") {
      
          my_formula <- paste0(Y, " ~ ",  rhs)
          
          m <- lm_robust(as.formula(my_formula), 
                         data = filter(data, points == j))
          rsq <- m$r.squared
          n <- m$N
          
          ## partial RSQ
          
          if (!bivariate) {
            f_reduced <- paste0(Y, " ~ ",  paste0(setdiff(XX, x), collapse = '+')) %>% 
              as.formula()
            f_full <- my_formula %>% as.formula()
            
            m_reduced <- lm(f_reduced, data = filter(data[data$points == j & 
                                                            complete.cases(data[, unique(c(XX, x))]), ]))
            m_full <- lm(f_full, data = filter(data[data$points == j & 
                                                      complete.cases(data[, unique(c(XX, x))]), ]))
            
            ## Get partial rsq
            
            rsq <- (sum(m_reduced$residuals^2) - sum(m_full$residuals^2)) / sum(m_reduced$residuals^2)
            
            ## Adjust: Formula is 1-(1-R2)*(DF_Reduced / DF_Full)
            
            rsq <- 1-((1-rsq)*(summary(m_reduced)$df[2] / summary(m_full)$df[2]))
          }
          
          ## Return
          
          m %>% tidy() %>% slice(2) %>% 
            mutate(n = n, rsq = rsq)
          
        } else {
          
          ## Negative binomial model
          my_formula <- paste0(Y, " ~ ",  rhs, '|', rhs)
  
          model <- zeroinfl(formula = as.formula(my_formula), data = data, dist = "negbin")
          
          out <- 
            rbind(
            summary(model)$coefficients$count[2,],
            summary(model)$coefficients$zero[2,]) %>% 
            data.frame() 
          
          names(out) <- c("estimate", "std.error", "statistic", "p.value")
          
          out %>%
            mutate(
              term = paste0(X, c("_count", "_zero")),
              conf.high  = estimate + 1.96*std.error,
              conf.low  = estimate - 1.96*std.error,
              outcome = "Y",
              n = model$n
              )
        }
          
      }) %>%
        bind_rows()  %>%                     
        select(term, estimate, conf.low, conf.high, n, rsq) %>% 
        mutate(elapsed = j) 
      
    }) %>%
    
      bind_rows() %>%
      data.frame()
    
    ## If labels are desired recode the term variable
    
    if (!is.na(Xlab[1])) {
      tempdf <- data.frame(term = X, label = Xlab)
      out <- left_join(out, tempdf) %>% 
        mutate(term = label)
    }
  
  ## Return 
  
  out %>% 
    group_by(term) %>% 
    arrange(elapsed) %>% 
    mutate(n = last(n), rsq = last(rsq)) %>% 
    mutate(label_nrsq = paste0('N=', n, ', ', 'R2=', round(rsq, 2))) %>% 
    ungroup()
}


do_plots <- function(ests, xlab)
  ests %>% 
  transform(term=factor(term,levels=unique(term))) %>%
    ggplot(aes(x=elapsed, y=estimate)) + 
      geom_point() +
      geom_line() +
      geom_errorbar(aes(ymin=conf.low, ymax=conf.high), colour="black", width=.1) +
      xlab(xlab) +
      ylab("Coefficient") +
      theme_bw()  +
    theme(panel.grid.minor = element_blank()) +
  geom_abline(slope = 0, intercept =0, linetype = 'dashed', color = "black") +
      facet_wrap(~term + label_nrsq, scales = "free")

# Get measures and labels


families <- c("state_cap_vars", "pol_account_vars", "social_vars",
              "econ_vars",  "phys_vars", "epi_vars", "health_sys_vars")

for(v in families){  
  assign(v, filter(measures, family == v)$vars %>% as.character) 
  assign(paste0(v, "_labels"), filter(measures, family == v)$labels %>% as.character)
 }

```

```{r, include = FALSE, eval = dographs, message = FALSE}
# Controls graphs, used in paper

pdf("figs/deaths_cum_log_controls.pdf", width = 8, height = 6)
  Y              <- "deaths_cum_log"
  relative_dates <- FALSE
  xlab           <- "Days since 1 Jan 2020"
  
  out <- results(Y = Y, X =  controls, Xlab = controls_labels,
          bivariate = FALSE, relative_dates = relative_dates)
   do_plots(out, xlab)
dev.off()


pdf("figs/deaths_cum_log_controls_biv.pdf", width = 8, height = 6)
  Y              <- "deaths_cum_log"
  relative_dates <- FALSE
  xlab           <- "Days since 1 Jan 2020"
  
  out <- results(Y = Y, X =  controls, Xlab = controls_labels,
          bivariate = TRUE, relative_dates = relative_dates)
   do_plots(out, xlab)
dev.off()
```

Existing social science research has implications for the kinds of features of  societies that might help explain variation in a society's ability to respond to adverse shocks. In some accounts, for example, more democratic societies will be more responsive; in other accounts, state capacity or social cohesion are more important. We gather together major measures suggested by these literatures and report on the evolving covariance between these measures and the cross national distribution of reported Covid-19 burdens. 

As of the most recent data date, `r format(data_date, '%d %B, %Y')`: 

```{r, echo = FALSE}
Mbase <- lm_robust(as.formula(paste("deaths_cum_log ~ ", paste(controls, collapse = "+"))),
           	data = filter(df, as.Date(df$date_rep) == data_date))
```



* A small set of simple controls---`r controls_text`---explain `r round(100*Mbase$r.squared, 0)`% of  the cross national variation in deaths (logged). We used a Lasso procedure to chose the controls that perform best at predicting Covid-19 mortality (see paper for more details).
* Given these controls:
    *  many political and economic measures are (currently) not related with Covid-19 burdens in the directions suggested by prominent theories. For instance, wealthier, more democratic, and more prepared states currently suffer the greatest burdens. 
    * A number of features highlighted in public discussions---such as the role of populist governments and women leaders---also do not find support. 
    * Other accounts do find support however, including a set of measures of state capacity and measures of social trust and trust in government.
* These patterns may change as the pandemic evolves.

* See [our paper](https://wzb-ipi.github.io/corona/WD_paper.pdf) discussing the literature and logics behind these analyses.
* See [this paper](https://wzb-ipi.github.io/corona/paper.pdf) for even more discussion as well as supplementary analysis on policy outcome measures.
* All data preparation code and sources available [here](https://wzb-ipi.github.io/corona/dataprep.html). And merged complete data [here](https://wzb-ipi.github.io/corona/df_full.csv).

**Notes / warnings**: 

* The covariances we report should **not be taken as estimates of causal effects** but rather as summaries of covariation between measures. 
* Cross national covariation can in part reflect **differences in reporting standards** and not just underlying variation in the disease.
* Currently outcome data is highly **skewed**. Moreover the **scale** of the outcome is changing as the disease spreads which in itself can produce larger coefficient magnitudes over time.

#  {.tabset}

## Overview {.tabset  .tabset-pills}

### Deaths/million (logged)

**Predicted and observed mortality**

---

Each point in the figure shows the actual and predicted (log) per capita mortality for a given country. To predict mortality, we use four covariates selected by a Lasso procedure that maximize predictive accuracy as well an additional control for the quality of health data reporting. The controls are `r controls_deaths_capita_text`.

```{r, message = FALSE, warning = FALSE, fig.cap = "\\label{oos2} Out of sample predictions from controls. Points above (below) the 45 degree line do worse (better) than predicted.", echo = FALSE}


oos_plot(depvar = "deaths_cum_per_million_log", 
         ylabel = "Actual Deaths per Million (logged)", 
         controls = controls_deaths_capita)


```

---


**Correlates of mortality**


We show conditional and unconditional associations between political and social characteristics and logged deaths, at three points in time. Data for explanatory variables are typically given for 2018 or 2019. In addition to bivariate relationships, we show associations controlling for the five controls variables listed above. The coefficients are from linear models using $z$-standardized independent variables.

```{r, echo = FALSE,message=F, warning=F}
Y_list   <- c('deaths_cum_per_million_log')
Y_labels <- c('Deaths/million')
wd_subfamily <- c("general_capacity",
                      "sector-specific capacity",
                      "centralization_decision",
                      "responsiveness",
                      "political_priorities",
                      "social_structures")

## Get results

res_apr <- results_list_wd_families(data = df %>% filter(as.Date(date_rep) == "2020-04-01"),
                 Y_list = Y_list,
                 Y_labels = Y_labels,
                 wd_subfamily=wd_subfamily,
                 controls_use=controls_use, break_line_family = T) %>%
  mutate(Y = "1 apr",
             Y_labels = "Deaths/million (logged) | 1 April")%>%
  rename(family_labels = wd_subfamily_labels)

res_may <- results_list_wd_families(data = df %>% filter(as.Date(date_rep) == "2020-05-31"),
                 Y_list = Y_list,
                 Y_labels = Y_labels,
                 wd_subfamily=wd_subfamily,
                 controls_use=controls_use, break_line_family = T) %>%
  mutate(Y = "31 may",
             Y_labels = "Deaths/million (logged) | 31 May")%>%
  rename(family_labels = wd_subfamily_labels)


res_today <- results_list_wd_families(data = df_today,
                 Y_list = Y_list,
                 Y_labels = Y_labels,
                 wd_subfamily = wd_subfamily,
                 controls_use = controls_use, break_line_family = T)%>%
  mutate(Y = "today",
             Y_labels = paste0("Deaths/million (logged) | ", format(data_date, '%d %B'))) %>%
  rename(family_labels = wd_subfamily_labels)


## Combine results


res_combined <- rbind(res_apr, res_may, res_today) %>% 
  mutate(Y_labels = factor(Y_labels, levels  = c("Deaths/million (logged) | 1 April",
                                                 "Deaths/million (logged) | 31 May",
                                                 paste0("Deaths/million (logged) | ", format(data_date, '%d %B')))))
```

```{r, echo = F, message=F, warning=F, comment=NA, error=FALSE, fig.width=8.5, fig.height=10, fig.cap = "\\label{results} Correlates of Covid-19 outcomes. Points represent normalized coefficients from either a bivariate model (circles) or a model with controls (squares).", fig.pos="!ht"}

plotresults(data = res_combined)
```



### Deaths total (logged) 

**Predicted and observed mortality**

Each point in the figure shows the actual and predicted (log) total mortality for a given country. To predict mortality, we use four covariates selected by a Lasso procedure that maximize predictive accuracy as well an additional control for the quality of health data reporting. The controls are `r controls_text`.

```{r, message = FALSE, warning = FALSE, fig.cap = "\\label{oos1} Out of sample predictions from controls. Points above (below) the 45 degree line do worse (better) than predicted.", echo = FALSE}


oos_plot(depvar = "deaths_cum_log", 
         ylabel = "Actual Deaths (logged)", 
         controls = controls)


```

---

**Correlates of mortality**

We show conditional and unconditional associations between political and social characteristics and logged deaths, at three points in time. Data for explanatory variables are typically given for 2018 or 2019. In addition to bivariate relationships, we show associations controlling for the five controls variables listed above. The coefficients are from linear models using $z$-standardized independent variables. 

```{r, echo = FALSE,message=F, warning=F}
Y_list   <- c('deaths_cum_log')
Y_labels <- c('Deaths/million')
wd_subfamily <- c("general_capacity",
                      "sector-specific capacity",
                      "centralization_decision",
                      "responsiveness",
                      "political_priorities",
                      "social_structures")

## Get results

res_apr <- results_list_wd_families(data = df %>% filter(as.Date(date_rep) == "2020-04-01"),
                 Y_list = Y_list,
                 Y_labels = Y_labels,
                 wd_subfamily=wd_subfamily,
                 controls_use=controls_use, break_line_family = T) %>%
  mutate(Y = "1 apr",
             Y_labels = "Total deaths (logged) | 1 April")%>%
  rename(family_labels = wd_subfamily_labels)

res_may <- results_list_wd_families(data = df %>% filter(as.Date(date_rep) == "2020-05-31"),
                 Y_list = Y_list,
                 Y_labels = Y_labels,
                 wd_subfamily=wd_subfamily,
                 controls_use=controls_use, break_line_family = T) %>%
  mutate(Y = "31 may",
             Y_labels = "Total deaths (logged) | 31 May")%>%
  rename(family_labels = wd_subfamily_labels)


res_today <- results_list_wd_families(data = df_today,
                 Y_list = Y_list,
                 Y_labels = Y_labels,
                 wd_subfamily = wd_subfamily,
                 controls_use=controls_use, break_line_family = T)%>%
  mutate(Y = "today",
             Y_labels = paste0("Total deaths (logged) | ", format(data_date, '%d %B'))) %>%
  rename(family_labels = wd_subfamily_labels)

## Combine results


res_combined <- rbind(res_apr, res_may, res_today) %>% 
  mutate(Y_labels = factor(Y_labels, levels  = c("Total deaths (logged) | 1 April",
                                                 "Total deaths (logged) | 31 May",
                                                 paste0("Total deaths (logged) | ", format(data_date, '%d %B')))))
```

```{r, echo = F, message=F, warning=F, comment=NA, error=FALSE, fig.width=8.5, fig.height=10, fig.cap = "\\label{results} Correlates of Covid-19 outcomes. Points represent normalized coefficients from either a bivariate model (circles) or a model with controls (squares).", fig.pos="!ht"}

plotresults(data = res_combined)
```


## Deaths total (logged)  {.tabset  .tabset-pills}

Correlates of total deaths recorded to date (logged), by day since 1 Jan 2020.

```{r}
Y              <- "deaths_cum_log"
Y_lab          <- "Deaths total (logged)"
relative_dates <- FALSE
xlab           <- "Days since 1 Jan 2020"
```

```{r, child = "6_report_child.Rmd"}
```




## Deaths/million (logged) {.tabset  .tabset-pills}

Correlates of total deaths per million population (logged), by day since 1 Jan 2020.

```{r}
dographs <- FALSE
Y              <- "deaths_cum_per_million_log"
relative_dates <- FALSE
xlab           <- "Days since 1 Jan 2020"
```

```{r, child = "6_report_child.Rmd"}
```


## Deaths  (since onset)  {.tabset  .tabset-pills}

Correlates of total deaths by day since country recoded at least 10 cases. Note that there are possible discontinuities in these graphs here as sample composition changes.  

```{r}
Y              <- "deaths_cum_log"
relative_dates <- TRUE
xlab           <- "Days since first 10 cases"
```

```{r, child = "6_report_child.Rmd"}

```


<!-- ## Lockdowns  {.tabset  .tabset-pills} -->

<!-- Correlates of enactment of lockdown measures, by day since 1 Jan 2020. -->


<!-- ```{r} -->
<!-- Y              <- "lockdown_bin" -->
<!-- relative_dates <- FALSE -->
<!-- xlab           <- "Days since 1 Jan 2020" -->
<!-- ``` -->

<!-- ```{r, child = "6_report_child.Rmd"} -->
<!-- ``` -->


## Overview plots  {.tabset  .tabset-pills}

### Cumulative deaths

Cumulative deaths, selected countries:

```{r, warning = FALSE}

view_set <- c("CHN", "KOR","ITA", "ESP", "USA",
              "GBR", "DEU", "CHE", "IRN", "IND", "BRA")

df_full %>% filter(geoid2 %in% view_set & elapsed_rel > -5)  %>%
  ggplot(aes(x=elapsed_rel, y=deaths_cum, color = geoid2)) + 
  geom_line() +
  xlab("Days since reported 10th case") +
  ylab("Deaths") +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  geom_dl(aes(label = geoid2), method = list(dl.trans(x = x + .2), "last.points")) + 
  theme(legend.position = "none")
```

Cumulative deaths, countries with per capita income below $2000:

```{r, warning = FALSE}

df_full %>% filter(gdp_pc < 2 & elapsed_rel > -5)  %>%
  ggplot(aes(x=elapsed_rel, y=deaths_cum, color = geoid2)) + 
  geom_line() +
  xlab("Days since reported 10th case | Poorer countries (< USD 2 K)") +
  ylab("Deaths") +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  geom_dl(aes(label = geoid2), method = list(dl.trans(x = x + .2), "last.points")) + 
  theme(legend.position = "none")
```


Cumulative deaths, selected countries, log scale.

```{r, warning = FALSE}

df_full %>% filter(geoid2 %in% view_set & elapsed_rel > -5)  %>%
  ggplot(aes(x=elapsed_rel, y=deaths_cum_log, color = geoid2)) + 
  geom_line() +
  xlab("Days since 10th reported case") +
  ylab("Deaths (logged)")  +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  geom_dl(aes(label = geoid2), method = list(dl.trans(x = x + .2), "last.points")) + 
  theme(legend.position = "none")
```


Cumulative deaths per million population, log scale.

```{r, warning = FALSE}

df_full %>% filter(geoid2 %in% view_set & elapsed_rel > -5)  %>%
  ggplot(aes(x=elapsed_rel, y=deaths_cum_per_million, color = geoid2)) + 
  geom_line() +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  xlab("Days since 10th reported case") +
  ylab("Deaths per million") +
  geom_dl(aes(label = geoid2), method = list(dl.trans(x = x + .2), "last.points")) + 
  theme(legend.position = "none")  +
  scale_y_continuous(trans='log10') 
```


By region.

```{r, echo = FALSE, warning = FALSE}

df_reg <- df_full  %>% 
  group_by(region, elapsed) %>% 
  summarize(region_deaths_cuml = sum(deaths_cum, na.rm = TRUE)) %>%
  group_by(forcats::fct_explicit_na(region)) %>%
  mutate(relative_start = ifelse(any(region_deaths_cuml >= 1),
                                 min(elapsed[region_deaths_cuml >= 1]), 
                                 NA),
         elapsed_rel = elapsed - relative_start) %>%
  ungroup() %>%
filter(elapsed_rel > -1 & !is.na(region))  

df_reg %>%
  ggplot(aes(x=elapsed_rel, y=region_deaths_cuml + 1, color = region)) + 
  geom_line() +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  xlab("Days since 1st death") +
  ylab("Cumulative deaths") +
  geom_dl(aes(label = region), method = list(dl.trans(x = x + .2), "last.points")) + 
  theme(legend.position = "none")+
  scale_y_continuous(trans='log10') + xlim(0, max(df_reg$elapsed_rel, na.rm=TRUE) + 10)


```

### Total deaths

```{r, fig.cap="Distribution of deaths (Wealthier and less wealthy countries)", echo = FALSE, fig.width = 10, warning = FALSE}

df_full %>% filter(as.Date(df_full$date_rep) == data_date & deaths_cum > 10) %>%
  mutate(group = if_else(gdp_pc > median(gdp_pc, na.rm = TRUE), 
                              paste0("GDP/capita > ",
                                     paste0(round(1000*round(median(gdp_pc, na.rm = TRUE)), 0),
                                            " USD PPP")), 
                              paste0("GDP/capita < ",
                                     paste0(round(1000*round(median(gdp_pc, na.rm = TRUE)), 0),
                                            " USD PPP")))) %>% 
  filter(!is.na(group)) %>%
  mutate(country = str_replace_all(country, "_", " ")) %>% 
  ggplot(aes(x = reorder(country, 
                         deaths_cum_log), 
             y = deaths_cum)) +   
  geom_bar(stat = "identity") +
  labs(x = "Country", 
       y = "Total deaths (log scale)") +
  theme_bw() +
    theme(panel.grid.minor = element_blank()) +
  facet_wrap(group ~ ., scales = "free_y", ncol = 2) +
  scale_y_continuous(trans='log10') +
  coord_flip()

```


```{r, fig.cap="Distribution of deaths (regions)", echo = FALSE, fig.width = 10, fig.height = 20, warning = FALSE}

df_full %>% filter(as.Date(df_full$date_rep) == data_date & deaths_cum > 1) %>%
  filter(!is.na(region)) %>%
  mutate(country = str_replace_all(country, "_", " ")) %>% 
  ggplot(aes(x = reorder(country, 
                         deaths_cum_log), 
             y = deaths_cum)) +   
  geom_bar(stat = "identity") +
  labs(x = "Country", 
       y = "Total deaths (log scale)") +
  theme_bw() +
    theme(panel.grid.minor = element_blank()) +
  facet_wrap(region ~ ., scales = "free_y", ncol = 2) +
  scale_y_continuous(trans='log10') +
  coord_flip()

```

### Growth rates deaths

7 day deaths growth rates, selected countries. Note log10 scale.

```{r, warning = FALSE}

df_full %>% filter(geoid2 %in% c("IND", "ITA", "BRA", 
                            "USA", "GBR", "DEU", "IRN") & elapsed_rel_d > 20)  %>%
  ggplot(aes(x=elapsed_rel_d, y=100*deaths_cum_g7, color = geoid2)) + 
  geom_line() +
  scale_y_continuous(trans='log10') +
  xlab("Days since 10th Death") +
  ylab("7 day growth in Deaths (percent; 100 = doubling)") +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  geom_dl(aes(label = geoid2), method = list(dl.trans(x = x + .2), "last.points")) + 
  theme(legend.position = "none")
```


There is a major reversal in relative growth rates. In early stages growth rates were low in Sub Saharan Africa and South Asia and high in Europe and North America, these rankings have since been reversed. Latin America stands out for sustained high growth rates.

```{r, fig.cap="Growth rates, regions (excl China))", echo = FALSE, message = FALSE, warning = FALSE}

df_reg <- df_full  %>% 
  group_by(region, elapsed) %>% 
  filter(geoid2 != "CHN") %>%
  summarize(region_deaths_cuml = sum(deaths_cum, na.rm = TRUE),
            region_deaths_cuml_l7 = sum(deaths_cum_l7, na.rm = TRUE),
            region_deaths_gr_l7 = (region_deaths_cuml/region_deaths_cuml_l7 - 1)) %>%
  group_by(forcats::fct_explicit_na(region)) %>%
  mutate(relative_start = ifelse(any(region_deaths_cuml >= 100),
                                 min(elapsed[region_deaths_cuml >= 100]), 
                                 NA),
         elapsed_rel = elapsed - relative_start) %>%
  ungroup() %>%
filter(elapsed_rel > -1 & !is.na(region))  

df_reg %>%
  ggplot(aes(x=elapsed_rel, y=100*region_deaths_gr_l7 + .00001, color = region)) + 
  geom_line() +
  scale_y_continuous(trans='log10') +
  xlab("Days since 100th Death") +
  ylab("7 day growth in Deaths (percent; 100 = doubling)") +
  theme_bw() +
  xlim(c(NA, max(df_reg$elapsed_rel) + 30)) + 
  theme(panel.grid.minor = element_blank()) +
  geom_dl(aes(label = region), method = list(dl.trans(x = x + .2), "last.points")) + 
  theme(legend.position = "none")

```

<!-- ### Death and lagged cases  -->

<!-- Relation between deaths and previously reported cases  (1, 7 and 14 days prior). -->

<!-- ```{r, fig.height=12, fig.width=9, warning = FALSE}   -->
<!-- j <- c("IRN", "CHN", "GBR", "KOR", "USA", "ITA", "ESP", "DEU", "TUR")  -->

<!-- df_full %>% filter((GeoId2 %in% j) & (elapsed_rel >=0))  %>% -->
<!--   ggplot(aes(elapsed_rel)) +  -->
<!--   geom_line(aes(y = deaths_cumul +1, colour = "deaths_cumul")) +  -->
<!--   geom_line(aes(y = cases_cumul_L14 +1, colour = "cases_cumul_L14")) + -->
<!--   geom_line(aes(y = cases_cumul_L7 +1, colour = "cases_cumul_L7")) + -->
<!--   geom_line(aes(y = cases_cumul_L1 +1, colour = "cases_cumul_L1")) + -->
<!--   facet_wrap(~ GeoId2, scales = "free")+ -->
<!--   scale_y_continuous(trans='log10') + -->
<!--   xlab("Days since 10th case") + -->
<!--   theme_bw() + -->
<!--   theme(panel.grid.minor = element_blank()) + -->
<!--   ylab("Cumulative deaths + 1") -->

<!-- ``` -->


## Browse data  {.tabset  .tabset-pills}

```{r, echo  = FALSE}
library(DT)

data_date <- max(as.Date(df$date_rep), na.rm = TRUE)
df_display <- df_full %>% filter(as.Date(date_rep) == data_date) %>% 
  select(geoid2, everything(), country) %>%
  select(-starts_with("X")) 
rownames(df_display) <- df_display$country
df_display %>% mutate_if(is.numeric, ~round(., 3)) %>%
datatable(caption = paste('Data for ', data_date), 
          rownames = TRUE, 
          extensions = c('Buttons', 'Scroller', 'ColReorder', 'FixedColumns'), 
          options = list(colreorder = TRUE, 
                         dom = 'Bfrtip',
                         buttons = c('copy', 'csv', 'excel'),
                         deferRender = TRUE,
                          scrollY = 200,
                          scroller = TRUE,
                         scrollX = TRUE
#,
#          fixedColumns = list(leftColumns = 1)
  )) 

```

