---
title: "graphs child"
author: "ipi"
date: "3/29/2020"
output: html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo = FALSE}
  if(Y == "deaths_cum_log") {
    XX <- controls
    Clab <- controls_labels

  } else {
    XX <- controls_deaths_capita
    Clab <- controls_deaths_capita_labels
  }

```  

### State capacity

```{r, fig.cap = control_text, warning = FALSE, message = FALSE}

control_text <- paste0("Controls: ", paste(Clab, collapse = ", "))

X = state_cap_vars
Xlab = state_cap_vars_labels

out <- 
  results(Y = Y, X =  X, Xlab = Xlab, XX = XX,
        bivariate = FALSE, relative_dates = relative_dates) 

do_plots(out, xlab)

```

```{r, include = FALSE, eval = dographs}
pdf(paste0("figs/", Y,"_state.pdf"), width = 8, height = 6)
 do_plots(out, xlab)
dev.off()

```

```{r, fig.cap = "No controls", warning = FALSE, message = FALSE}
results(Y = Y, X =  X, Xlab = Xlab, XX=XX,
        bivariate = TRUE, relative_dates = relative_dates) %>% do_plots(xlab)
```


```{r, include = TRUE, eval = TRUE}
filter(measures, family == "state_cap_vars"  | family == "outcome") %>% select(labels, Source, Definition) %>% kable(caption = "Sources")
```

```{r, child = "7_report_child_notes.Rmd"}

```

### Political Institutions

```{r, fig.cap = control_text, warning = FALSE, message = FALSE}
X = pol_account_vars
Xlab = pol_account_vars_labels

out <- 
  results(Y = Y, X =  X, Xlab = Xlab, XX=XX,
        bivariate = FALSE, relative_dates = relative_dates) 

do_plots(out, xlab)

```


```{r, include = FALSE, eval = dographs}
pdf(paste0("figs/", Y,"_account.pdf"), width = 8, height = 6)
 do_plots(out, xlab)
dev.off()

```

```{r, fig.cap = "No controls", warning = FALSE, message = FALSE}
results(Y = Y, X =  X, Xlab = Xlab, XX=XX,
        bivariate = TRUE, relative_dates = relative_dates) %>% do_plots(xlab)

filter(measures, family == "pol_account_vars" | family == "outcome") %>% select(labels, Source, Definition) %>% kable(caption = "Sources")

```



```{r, child = "7_report_child_notes.Rmd"}

```



### Social structures

```{r, fig.cap = control_text, warning = FALSE, message = FALSE}
X = social_vars
Xlab = social_vars_labels

out <- results(Y = Y, X =  X, Xlab = Xlab, XX=XX,
        bivariate = FALSE, relative_dates = relative_dates)

do_plots(out, xlab)
```


```{r, include = FALSE, eval = dographs}

pdf(paste0("figs/", Y,"_social.pdf"), width = 8, height = 6)
 do_plots(out, xlab)
dev.off()
```

```{r, fig.cap = "No controls", warning = FALSE, message = FALSE}
results(Y = Y, X =  X, Xlab = Xlab, XX=XX,
        bivariate = TRUE, relative_dates = relative_dates) %>% do_plots(xlab)
```



```{r}
filter(measures, family == "social_vars"  | family == "outcome") %>% select(labels, Source, Definition) %>% kable(caption = "Sources")
```



```{r, child = "7_report_child_notes.Rmd"}

```



### Global Linkages

```{r, fig.cap = control_text, warning = FALSE, message = FALSE}
X = econ_vars
Xlab = econ_vars_labels

out <- results(Y = Y, X =  X, Xlab = Xlab, XX=XX,
        bivariate = FALSE, relative_dates = relative_dates)

do_plots(out, xlab)
```


```{r, include = FALSE, eval = dographs}

pdf(paste0("figs/", Y,"_econ.pdf"), width = 8, height = 6)
 do_plots(out, xlab)
dev.off()
```

```{r, fig.cap = "No controls", warning = FALSE, message = FALSE}
results(Y = Y, X =  X, Xlab = Xlab, XX=XX,
        bivariate = TRUE, relative_dates = relative_dates) %>% do_plots(xlab)
```


```{r}
filter(measures, family == "econ_vars"  | family == "outcome") %>% select(labels, Source, Definition) %>% kable(caption = "Sources")
```



```{r, child = "7_report_child_notes.Rmd"}

```



### Environmental risks


```{r, fig.cap = control_text, warning = FALSE, message = FALSE}
X = phys_vars
Xlab = phys_vars_labels

out <- results(Y = Y, X =  X, Xlab = Xlab, XX=XX,
        bivariate = FALSE, relative_dates = relative_dates)

do_plots(out, xlab)
```


```{r, include = FALSE, eval = dographs}
pdf(paste0("figs/", Y,"_phys.pdf"), width = 8, height = 6)
 do_plots(out, xlab)
dev.off()
```

```{r, fig.cap = "No controls", warning = FALSE, message = FALSE}
results(Y = Y, X =  X, Xlab = Xlab, XX=XX,
        bivariate = TRUE, relative_dates = relative_dates) %>% do_plots(xlab)
```


```{r}
filter(measures, family == "phys_vars"  | family == "outcome") %>% select(labels, Source, Definition) %>% kable(caption = "Sources")
```


```{r, child = "7_report_child_notes.Rmd"}

```


### Health risks


```{r, fig.cap = control_text, warning = FALSE, message = FALSE}
X = epi_vars
Xlab = epi_vars_labels

out <- results(Y = Y, X =  X, Xlab = Xlab, XX=XX,
        bivariate = FALSE, relative_dates = relative_dates)

do_plots(out, xlab)
```


```{r, include = FALSE, eval = dographs}
pdf(paste0("figs/", Y,"_epi.pdf"), width = 8, height = 6)
 do_plots(out, xlab)
dev.off()
```

```{r, fig.cap = "No controls", warning = FALSE, message = FALSE}
results(Y = Y, X =  X, Xlab = Xlab, XX=XX,
        bivariate = TRUE, relative_dates = relative_dates) %>% do_plots(xlab)
```


```{r}
filter(measures, family == "epi_vars"  | family == "outcome") %>% select(labels, Source, Definition) %>% kable(caption = "Sources")
```


```{r, child = "7_report_child_notes.Rmd"}

```


### Health system


```{r, fig.cap = control_text, warning = FALSE, message = FALSE}
X = health_sys_vars
Xlab = health_sys_vars_labels

out <- results(Y = Y, X =  X, Xlab = Xlab, XX=XX,
        bivariate = FALSE, relative_dates = relative_dates)

do_plots(out, xlab)

```


```{r, include = FALSE, eval = dographs}
pdf(paste0("figs/", Y,"_health_sys.pdf"), width = 8, height = 6)
 do_plots(out, xlab)
dev.off()
```

```{r, fig.cap = "No controls", warning = FALSE, message = FALSE}
results(Y = Y, X =  X, Xlab = Xlab, XX=XX,
        bivariate = TRUE, relative_dates = relative_dates) %>% do_plots(xlab)
```


```{r, eval = TRUE, include = TRUE}
filter(measures, family == "health_sys_vars"  | family == "outcome") %>% select(labels, Source, Definition) %>% kable(caption = "Sources")
```



```{r, child = "7_report_child_notes.Rmd"}

```



### Controls  


```{r, echo = FALSE}
if(Y == "deaths_cum_log") {
    X <- controls
    Xlab <- controls_labels}

  if(Y == "deaths_cum_per_million_log") {
    X <- controls_deaths_capita
    Xlab <- controls_deaths_capita_labels}

Mbase2 <- lm_robust(as.formula(paste(Y, " ~ ", paste(X, collapse = "+"))),
           	data = filter(df, as.Date(df$date_rep) == data_date))

```

As of `r format(data_date, '%d %B, %Y')`, the five control variables account for `r round(100*Mbase2$r.squared, 2)`% of  the cross national variation in deaths.

The $R^2$ in the figures below shows the share of the variation that is explained by each control variable on its own. 

```{r, fig.cap = "Bivariate relations between controls and deaths (logged)", echo = FALSE, message = FALSE, warning = FALSE}

  relative_dates <- FALSE
  xlab           <- "Days since 1 Jan 2020"
  
  out <- results(Y = Y, X =  X, Xlab = Xlab, XX=XX,
          bivariate = TRUE, relative_dates = relative_dates)
   do_plots(out, xlab)
```

The $R^2$ in the figures below shows the share of the residual variation that is explained by each control variable after taking account of the other control variables.

```{r, fig.cap = "Relations between controls and deaths (logged), conditional on other controls", echo = FALSE, message = FALSE, warning = FALSE}
  relative_dates <- FALSE
  xlab           <- "Days since 1 Jan 2020"
  
  out <- results(Y = Y, X =  X, Xlab = Xlab, XX = XX,
          bivariate = FALSE, relative_dates = relative_dates)
   do_plots(out, xlab)
```
